---
title: "DEG group comparison analysis for Bladder Cancer - Stage 1 vs. Pre Cancer"
Author: "Tazeen Shaukat"
output:
  pdf_document: default
  html_notebook: default
---

# Project: Bladder Cancer Data Analysis

#### Purpose: Exploring molecular changes and identify potential bio-markers at different stages of bladder cancer, as well as differences in immune cells types that have infiltrated tumors.

#### Team 4: My team is tasked with completing two comparisons. These include comparing non invasive stage vs. precancerous tissue, and invasive stage vs non invasive stage cancer tissue

#### [**This file: Differential Gene Expression analysis using Ttest in R (group comparison) for Part I - Comparing Stage 1 (non-invasive) Cancer vs. Precancerous Tissue.**]{.underline}

### 1a - Read in clinical data

```{r}
clinData <- read.csv(file = "input/BC_ClinData_233rows.csv", header = T, stringsAsFactors = F, row.names = 1)


head(clinData)
knitr::kable(head(clinData), caption = "Clinical Data with rows as patients and attributes as columns")
```

### Checking the shape and overall information about the clinical dataset

```{r}
dim(clinData)

```

### 1b: Read in processed normalized gene expression data in log2 scale, includes gene annotation

```{r}

# Read in gene expression file - features (rows), 30 patients (columns)
geneExp <- read.csv(file = "input/BC_GeneExpData_withAnno_233.tsv", sep = "\t", header = T, stringsAsFactors = F, row.names = 1)
head(geneExp[1:5, 1:4])
knitr::kable(head(geneExp), caption = "Processed gene expression data in log2 scale with gene annotation")
```

### Now lets combine the dataset using the Patient's GSMID as the unique identifier

```{r}
head(clinData$GSMid) # these are the identifiers in the clinical data 
head(colnames(geneExp)) # same identifiers in the gene exp data 

matchingSamples = which(colnames(geneExp) %in% (clinData$GSMid)) # 233 samples
head(matchingSamples)
subsetGeneExp <- geneExp[, matchingSamples]
head(subsetGeneExp)
```

## Step 3 - Identifying the groups to be compared

### Identifying the groups to be compared (Baseline and Comparison Grps)

In this case:

-   Baseline = Bladder mucosae surrounding cancer (Precancerous)

-   Comparison = Primary_BC_Superficial (Stage 1 Non Invasive)

```{r}
# Labels (row numbers) that can identify the baseline group patients
baselineGrpLabels <-  which(clinData$PrimaryBladderCancerType == "Bladder mucosae surrounding cancer") #58 samples
head(baselineGrpLabels)
length(baselineGrpLabels)

# Use the labels (row numbers) to subset baseline patients in clinical data file 
clinBase <- clinData[baselineGrpLabels, ]
head(clinBase)

# Labels (row numbers) that can identify the comp group patients
compGrpLabels <- which(clinData$PrimaryBladderCancerType == "Primary_BC_Superficial") #103 samples
head(compGrpLabels)
length(compGrpLabels)

# Use the labels (row numbers) to subset comp patients in clinical data file 
clinComp <- clinData[compGrpLabels, ]
head(clinComp)

#### Use the clinBase and clinComp objects to subset gene expression data
geneExpBase <- subsetGeneExp[, clinBase$GSMid] # 43148 feature (rows), 58 samples columns
geneExpComp <- subsetGeneExp[, clinComp$GSMid] # 43148 feature (rows), 103 samples columns

head(geneExpBase)
head(geneExpComp)
```

## Step 4: Sanity check

-   See if filtering of clinical data in R matches filtering of clinical data in excel
-   See if sample ids in clinical data match sample ids in gene exp data (if they don't match it means your step 1 and/or 2 is wrong)
-   Verify you see correct number of samples in baseline and comp groups
-   Export the column names from gene expression data to see if it contains only probe/gene names and no other garbage

```{r}
#See if sample ids in clinical data match sample ids in gene exp data
clinBase$GSMid == colnames(geneExpBase)
clinComp$GSMid == colnames(geneExpComp)

### Checking if the samples IDs baseline and comp groups are the same ---- can put these in an excel to see if IDs match
write.table(clinBase$GSMid , file = "sanity/Tazeen_ClinBaseIDs.tsv", sep="\t", quote = F  )
write.table(clinComp$GSMid , file = "sanity/Tazeen_ClinCompIDs.tsv", sep="\t", quote = F  )
write.table(colnames(geneExpBase) , 
            file = "sanity/Tazeen_GeneExpBaseIDs.tsv", 
            sep="\t", quote = F  )
write.table(colnames(geneExpComp) , 
            file = "sanity/Tazeen_GeneExpCompIDs.tsv", 
            sep="\t", quote = F  )

#Export the features from gene expression data  
#Open this file and check that it contains only probe/gene names and no other garbage
write.table(rownames(geneExp),file = "Tazeen_FeatureIDs.tsv", sep="\t", quote = F  )
```

## Step 5: Preparing data for T-test

-   Molecular data must have features (genes in this case) as rows, and samples as columns.
-   Transpose data (if needed) to obtain this
-   Objects must be data frame
-   Numeric data only

```{r}
### Checking to make sure data is a numeric data frame
knitr::kable(head(geneExpBase[1:5,1:4]))
knitr::kable(head(geneExpComp[1:5,1:4]))
```

#### Function for T-test

```{r ttest}
source("fnTTest.R")

#### Call T-test function
results1 = fnTTest(baseGroup = geneExpBase, 
                   compGroup = geneExpComp, 
                   testName = "Tazeen_Team4_Step1_TTest_", 
                   baseGroupName = "Precancer",
                   compGroupName =  "Stage1NonInvasive",
                   folderName = "output")
```

## Final Step - Sub-set top differentially expressed genes

```{r}
#Read in the T-Test results file

ttestResults <- read.csv(file = "output/Tazeen_Team4_Step1_TTest__Stage1NonInvasive_(Comp).vs._Precancer_(Base).TTest.csv")

#check to make sure p-value column is imported as numeric 
#sort by p-value (just in case the results are not sorted by p-value)

ttestResultsSorted <- dplyr::arrange(ttestResults, Pvalue)

#find rows with p-value < 0.05
whichSig <- which(ttestResultsSorted$Pvalue <= 0.05) 

#Short list sig results
ttestResultsSig <- ttestResultsSorted[whichSig, ] #18395 rows 

### Export short listed results
write.table(x = ttestResultsSig, 
            file = "output/Tazeen_Stage1_Precancerous_Ttest_Shortlisted.csv", 
            quote = F, sep = ",")

##### First column is a list of features in thsi format : ProbeID|GeneName. 
#### Use string split strsplit() function to extract gene names
funcSplit <- function(featureX) {
  f1 <- unlist(strsplit(x = featureX, split = "|", fixed = TRUE))
  f2 <- f1[2]
  return(f2)
}

# Use apply() function to run the split on every row, its faster version of a loop
geneNames1 <- apply(X = as.matrix(ttestResultsSig$Feature), 
                    MARGIN = 1, FUN = funcSplit)

head(geneNames1)

#print length of short listed gene names
length(geneNames1)

### Export list of gene names
write.table(x = geneNames1, 
            file = "output/Tazeen_Stage1_Precancerous_SigDiffExpressedGenes.csv", 
            quote = F, sep = ",")
```
