---
title: "Enrichment Analysis for Bladder Cancer: Stage2 vs Stage1"
author: "Tazeen Shaukat"
output:
  pdf_document: default
  html_document: default
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#install.packages("enrichR")
#install.packages("openxlsx")# independant of java
#install.packages("devtools") 
#devtools::install_github("wjawaid/enrichR")
library(enrichR)
library(openxlsx)

# NOTE - if you have an error "EnrichR website not responding"  and unable to connect to EnrichR database online, try this
# OPTION 1
#devtools::install_github("wjawaid/enrichR")
#library(enrichR)

# OPTION 2 - update curl and wget
# (Step a) Open the "terminal" application for these commands. Follow instructions for mac
# https://help.ubidots.com/en/articles/2165289-learn-how-to-install-run-curl-on-windows-macosx-linux
# 
# (Step b) Install/update wget to the latest version
# Use the same "terminal" application. Enter
# brew install wget
# 
# (step C) Restart your computer. 
# Try out these EnrichR commands to see if you are able to access the enrichR databases online without error message
# install.packages("enrichR")
# library(enrichR)
# enrichR::listEnrichrDbs()

```

## R Markdown - Enrichment anlalysis using EnrichR R package

Summary

-   I have already shortlisted the genes in my ...\_Step1.rmd file based
    on p-value cut off of 0.05

-   I also have created a clean list of gene names in the same file
    above

-   However, I will be redoing the filtering and gene name clean up to
    ensure I get the same results, and also to make sure the gene names
    are given to the EnrichR function in the correct/expected format

-   After discussing with my group, I will also be adding another layer
    of filtering the results on fdr cut off of 0.01 and limit the result
    set to a list of no more than a 1000 genes

-   Then I will connect to the EnrichR database and run Enrichment

## Import T-test results so that we can short list further

```{r}
#folder that contains group comparison results
fileName <- "input/Tazeen_Stage2_Stage1_Ttest_Shortlisted.csv"

# read in group comparison results
results <- read.table(file = fileName, 
                      header = TRUE, 
                      stringsAsFactors = FALSE, 
                      row.names = 1, 
                      sep = ",")
```

## Filtering/shortlisting Short list results based in p-value cut off (pvalue \<= 0.01) This is already done but redoing just to be sure

```{r}
pValueCutOff <- 0.05
fdrCutOff <- 0.01
which <- (as.numeric(results$Pvalue) <= pValueCutOff) & (as.numeric(results$FDR) <= fdrCutOff)
table(which) #10893 genes with p <= 0.05 and fdr <= 0.01


resultShort <- results[which, ] #short listed genes 
resultShort <- head(resultShort[order(as.numeric(resultShort$FDR)), ], 1000) #keeping only the top 1000 based on fdr

```

## Clean gene names

Note - the gene names are in the form of "reporterid_genename". So need
to split this.

```{r}
funcSplit <- function(rep_gene) {
    rep_gene_split <- unlist(strsplit(x = rep_gene, 
                                      split = "|", 
                                      fixed = TRUE))
    gene <- rep_gene_split[2]
    return(gene)
}
geneListSplit <- apply(X = as.matrix((resultShort$Feature)), 
                       MARGIN = 1, FUN = funcSplit )
head(geneListSplit) #cleaned gene names

#remove duplicates
geneListSplit_unique <- unique(geneListSplit)

# remove NA value
geneList_final <- na.omit(geneListSplit_unique) 

head(geneList_final)
```

```{r}
#print number of unique genes
# left with 950 after cleanup
length(geneList_final)
```

## Load Databases for Enrichr R package , and check connection

```{r}
#checking if EnrichR website and packing are working
#dbs <- enrichR::listEnrichrDbs()  #total number of databases available = 200+ 

#testing if EnrichR package is working
testOutput <- enrichR::enrichr(genes = c("Runx1", "Gfi1", "Gfi1b", "Spi1", "Gata1", "Kdr"), databases = "KEGG_2021_Human")
head(testOutput[[1]])

#List of databases for which enrichment analysis will be run
dblist1 <- read.csv(file = "input/2023-EnrichR-Databases.txt", 
                    header = F, stringsAsFactors = F)

head(dblist1)
```

```{r}
geneList_final_df <- data.frame(Gene = geneList_final)
head(geneList_final_df)

```

## Call function to run Enrichment

```{r, message=FALSE, warning=FALSE}
# set output file name
outputFileName <- paste("output/Stage2vsStage1.", "_EnrichR.xlsx", sep="")

#Load R script into the environment
source(file = "functionEnrichment.R") 

#call function to run Enrichment
functionEnrichment(dblist1, geneList_final_df, outputFileName)

#NEED INTERNET CONNECTION
```

Note - you will need internet connection to complete the above step.
